---
title: "Cyclistic bike-share analysis case study"
output: html_document
date: "2023-08-10"
---


#### Load Packages

```{r Installing and loading packages}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("dplyr")
install.packages("ggplot2")

library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
```

#### Import data and store it appropriately

For this case study I chose 12 data files containing trip data gathered during each month of 2022. Data has been downloaded from [Motivate International Inc.](https://divvy-tripdata.s3.amazonaws.com/index.html) Local copies have been stored securely on Google Drive and here on Kaggle.

```{r Gather data}
y2022_01 <- read_csv("Bike data /202201-divvy-tripdata.csv")
y2022_02 <- read_csv("Bike data /202202-divvy-tripdata.csv")
y2022_03 <- read_csv("Bike data /202203-divvy-tripdata.csv")
y2022_04 <- read_csv("Bike data /202204-divvy-tripdata.csv")
y2022_05 <- read_csv("Bike data /202205-divvy-tripdata.csv")
y2022_06 <- read_csv("Bike data /202206-divvy-tripdata.csv")
y2022_07 <- read_csv("Bike data /202207-divvy-tripdata.csv")
y2022_08 <- read_csv("Bike data /202208-divvy-tripdata.csv")
y2022_09 <- read_csv("Bike data /202209-divvy-tripdata.csv")
y2022_10 <- read_csv("Bike data /202210-divvy-tripdata.csv")
y2022_11 <- read_csv("Bike data /202211-divvy-tripdata.csv")
y2022_12 <- read_csv("Bike data /202212-divvy-tripdata.csv")
```

#### **Identify how data is organized.**

All trip data is in comma-delimited (.CSV) format with 13 columns, including: ride ID #, ride type, start/end time, starting point (id, name, and latitude/longitude), ending point (code, name, and latitude/longitude), and member/casual rider.

**Determine the credibility of the data.**

Due to the fact that this is a case study using public data, we are going to assume the data is credible.

## Process:

#### **Check for column consistency**

Use `colnames()` on each new data frame to make sure all have the same 13 columns.

```{r}
colnames(y2022_01)
colnames(y2022_02)
colnames(y2022_03)
colnames(y2022_04)
colnames(y2022_05)
colnames(y2022_06)
colnames(y2022_07)
colnames(y2022_08)
colnames(y2022_09)
colnames(y2022_10)
colnames(y2022_11)
colnames(y2022_12)
```

#### Combining data

`bind_rows()`: Combine rows from 12 different data frames into one large.

```{r}
Bike_rides_2022<- bind_rows (y2022_01, y2022_02, y2022_03, y2022_04, y2022_05, y2022_06, y2022_07, y2022_08, y2022_09, y2022_10, y2022_11, y2022_12)
```

#### Cleaning data

#Renaming columns

```{r}
Bike_rides_2022<- rename(Bike_rides_2022, "bike_type"="rideable_type", "user_type"="member_casual")
```

#Removing any unnecessary columns (erroneous latitude/longitude fields).

```{r}
Bike_rides_2022 <- Bike_rides_2022 %>% 
  select(-c( start_lat, start_lng, end_lat, end_lng))
```

#Removing rows with any missing values (NA) from a data frame

```{r}
Bike_rides_2022 <- na.omit(Bike_rides_2022)
```

#Ensuring datetime format is consistent throughout the started_at and ended_at columns.

```{r}
Bike_rides_2022$started_at <- ymd_hms(Bike_rides_2022$started_at)
Bike_rides_2022$ended_at <- ymd_hms(Bike_rides_2022$ended_at)
```

## Analyze

#### **Adding column for each ride length in minutes**.

#Using the `difftime()` function to calculate the length of each trip in minutes, rounded to two decimals.

```{r}
Bike_rides_2022$ride_length_min <- round(as.numeric(difftime(Bike_rides_2022$ended_at,Bike_rides_2022$started_at, units = "mins")), 2)

```

#### Filtering out data

Filtering out all null values and remove rides less than 60 seconds (or 1 minute) and greater than 24 hrs (or 1440 minutes) in length.

*After a thorough review of the data, I'll remove the following:*

-   Rides less than 60 seconds in length as they are potentially false starts or users trying to re-dock a bike to ensure it was secure per Divvy website: Divvy System Data
-   Rides with a negative ride_length_min are considered invalid since the trip start time cannot be greater than the trip end time
-   Rides with a ride_length_min greater than 24 hrs are considered invalid outliers for purposes of this project

```{r}
Bike_rides_2022<- Bike_rides_2022 %>%
  filter(!is.na(ride_length_min) & ride_length_min >= 1 & ride_length_min <= 1440)
```

#### **Adding columns for hour, time_of_day, day_of_week, month, and season to compare casual users and members**

#Adding column for hour of the day

This new column will contain the hour component extracted from the started_at column using the `lubridate::hour()` function.

```{r}
Bike_rides_2022$hour <- lubridate::hour(Bike_rides_2022$started_at)
```

#Adding column for time_of_day

```{r}
Bike_rides_2022 <- Bike_rides_2022 %>% mutate(time_of_day = case_when(
  hour >= 6 & hour < 9 ~ "Early Morning",
  hour >= 9 & hour < 12 ~ "Mid Morning",
  hour >= 12 & hour < 18  ~ "Afternoon",
  hour >= 18 & hour <= 23  ~ "Evening",
  hour >= 0 & hour < 3  ~ "Early Night",
  hour >= 3 & hour < 6  ~ "Late Night"))
```

#Adding a column for day_of_week the ride took place

```{r}
Bike_rides_2022<- Bike_rides_2022 %>%
  mutate(day_of_week = weekdays(started_at))
```

#Adding columns for the season and month

Function **`get_season`** is used to categorize months into seasons.

The **`sapply`** function is then used to apply the **`get_season`** function to each element in the **`month`** column.

```{r}
 get_season <- function(month) {
  if (month %in% c(12, 1, 2)) {
    return("Winter")
  } else if (month %in% c(3, 4, 5)) {
    return("Spring")
  } else if (month %in% c(6, 7, 8)) {
    return("Summer")
  } else {
    return("Fall")
  }
}

Bike_rides_2022 <- Bike_rides_2022 %>%
  mutate(month = as.integer(format(started_at, "%m")),
         season = sapply(month, get_season))
```

#### **Descriptive analysis on all rides (all figures in minutes)**

```{r}
mean(Bike_rides_2022$ride_length_min) #straight average (total ride length / rides)
median(Bike_rides_2022$ride_length_min) #midpoint number in the ascending array of ride lengths
max(Bike_rides_2022$ride_length_min) #longest ride
min(Bike_rides_2022$ride_length_min) #shortest ride
```

Key insights:

-   Average ride length is 17.27 minutes

-   Median ride length is 10.8 minutes

-   Maximum ride length is 1439.37 minutes (23.99 hours)

-   Minimum ride length is 1 min

#### **Casual rider vs the member rider**

```{r}
user_types_breakdown<-Bike_rides_2022 %>% 
  group_by(user_type) %>% 
  summarise(count = n(), Percentage = n()/nrow(Bike_rides_2022)*100)
```

``` r
# A tibble: 2 × 3
  user_type   count Percentage
  <chr>       <int>      <dbl>
1 casual    1731071       40.3
2 member    2561442       59.7
```

```{r}
ggplot(user_types_breakdown, aes(x = user_type, y = Percentage, fill = user_type)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(x = "User Type", y = "Percentage (%)", title = "Bike Ride Distribution by User Type") +
  scale_fill_discrete(name = "User Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Key insight:

-   59.7% of riders are members, and 40.3% of riders are casual users.

#### **Descriptive analysis on casual riders and members.**

Using `aggregate()` function to compare mean, median, max and min of casual riders and members

```{r}
aggregate(Bike_rides_2022$ride_length_min ~ Bike_rides_2022$user_type, FUN = mean)
aggregate(Bike_rides_2022$ride_length_min ~ Bike_rides_2022$user_type, FUN = median)
aggregate(Bike_rides_2022$ride_length_min ~ Bike_rides_2022$user_type, FUN = max)
aggregate(Bike_rides_2022$ride_length_min ~ Bike_rides_2022$user_type, FUN = min)

```

I found out that :

-   The mean ride length for casual users and members is 24.07 min and 12.67 min, respectively.

-   The median ride length for casual users and members is 14.08 min and 9.15 min, respectively.

-   Max and min ride lengths for both type of users were nearly the same

## **Further Analysis and Visualizations:**

#### **Adding a column `ride_length_cat` for categories to get a better visual in R.**

```{r}
Bike_rides_2022 <- Bike_rides_2022 %>% 
  mutate(ride_length_cat = case_when(
   ride_length_min <11.99 ~ "< 12 min",
   ride_length_min >=12 & ride_length_min <=20.99 ~ "12-20 min",
   ride_length_min >=21 & ride_length_min <=30.99  ~ "21-30 min",
   ride_length_min >=31 & ride_length_min <=60.99  ~ "31-60 min",  ride_length_min >=60 & ride_length_min <=120.99  ~ "61-120 min",
 ride_length_min >=121 & ride_length_min <=240.99  ~ "121-240 min",
 ride_length_min >=241  ~ "241+ min"))
```

#### **Creating a data frame that will show the percentage of rides for each `user_type` within each `ride_length_cat` category.**

I will use percentage instead of raw counts.

```{r}
summary_data <-Bike_rides_2022 %>% 
  group_by(user_type, ride_length_cat) %>%
  summarise(count = n()) %>%
  group_by(user_type) %>%
  mutate(total_count = sum(count)) %>%
  mutate(percentage = (count / total_count) * 100) %>%
  select(user_type, ride_length_cat, percentage) %>%
  ungroup()
```






