---
title: "Cyclistic bike-share analysis case study"
output: html_document
date: "2023-08-10"
---

## Introduction:

This analysis is for case study "Cyclistic bike-share analysis case study" from the Google Data Analytics Certificate.

## Ask:

#### **Identify the business task**

#### **Consider key stakeholders**

## Prepare:

#### Load Packages

```{r Installing and loading packages}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("dplyr")
install.packages("ggplot2")

library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
```

#### Import data and store it appropriately

For this case study I chose 12 data files containing trip data gathered during each month of 2022. Data has been downloaded from [Motivate International Inc.](https://divvy-tripdata.s3.amazonaws.com/index.html) Local copies have been stored securely on Google Drive and here on Kaggle.

```{r Gather data}
y2022_01 <- read_csv("Bike data /202201-divvy-tripdata.csv")
y2022_02 <- read_csv("Bike data /202202-divvy-tripdata.csv")
y2022_03 <- read_csv("Bike data /202203-divvy-tripdata.csv")
y2022_04 <- read_csv("Bike data /202204-divvy-tripdata.csv")
y2022_05 <- read_csv("Bike data /202205-divvy-tripdata.csv")
y2022_06 <- read_csv("Bike data /202206-divvy-tripdata.csv")
y2022_07 <- read_csv("Bike data /202207-divvy-tripdata.csv")
y2022_08 <- read_csv("Bike data /202208-divvy-tripdata.csv")
y2022_09 <- read_csv("Bike data /202209-divvy-tripdata.csv")
y2022_10 <- read_csv("Bike data /202210-divvy-tripdata.csv")
y2022_11 <- read_csv("Bike data /202211-divvy-tripdata.csv")
y2022_12 <- read_csv("Bike data /202212-divvy-tripdata.csv")
```

#### **Identify how data is organized.**

All trip data is in comma-delimited (.CSV) format with 13 columns, including: ride ID #, ride type, start/end time, starting point (id, name, and latitude/longitude), ending point (code, name, and latitude/longitude), and member/casual rider.

**Determine the credibility of the data.**

Due to the fact that this is a case study using public data, we are going to assume the data is credible.

## Process:

#### **Check for column consistency**

Use `colnames()` on each new data frame to make sure all have the same 13 columns.

```{r}
colnames(y2022_01)
colnames(y2022_02)
colnames(y2022_03)
colnames(y2022_04)
colnames(y2022_05)
colnames(y2022_06)
colnames(y2022_07)
colnames(y2022_08)
colnames(y2022_09)
colnames(y2022_10)
colnames(y2022_11)
colnames(y2022_12)
```

#### Combining data

`bind_rows()`: Combine rows from 12 different data frames into one large.

```{r}
Bike_rides_2022<- bind_rows (y2022_01, y2022_02, y2022_03, y2022_04, y2022_05, y2022_06, y2022_07, y2022_08, y2022_09, y2022_10, y2022_11, y2022_12)
```

#### Cleaning data

#Renaming columns

```{r}
Bike_rides_2022<- rename(Bike_rides_2022, "bike_type"="rideable_type", "user_type"="member_casual")
```

#Removing any unnecessary columns (erroneous latitude/longitude fields).

```{r}
Bike_rides_2022 <- Bike_rides_2022 %>% 
  select(-c( start_lat, start_lng, end_lat, end_lng))
```

#Removing rows with any missing values (NA) from a data frame

```{r}
Bike_rides_2022 <- na.omit(Bike_rides_2022)
```

#Ensuring datetime format is consistent throughout the started_at and ended_at columns.

```{r}
Bike_rides_2022$started_at <- ymd_hms(Bike_rides_2022$started_at)
Bike_rides_2022$ended_at <- ymd_hms(Bike_rides_2022$ended_at)
```

## Analyze

#### **Adding column for each ride length in minutes**.

#Using the `difftime()` function to calculate the length of each trip in minutes, rounded to two decimals.

```{r}
Bike_rides_2022$ride_length_min <- round(as.numeric(difftime(Bike_rides_2022$ended_at,Bike_rides_2022$started_at, units = "mins")), 2)

```

#### Filtering out data

Filtering out all null values and remove rides less than 60 seconds (or 1 minute) and greater than 24 hrs (or 1440 minutes) in length.

*After a thorough review of the data, I'll remove the following:*

-   Rides less than 60 seconds in length as they are potentially false starts or users trying to re-dock a bike to ensure it was secure per Divvy website: Divvy System Data
-   Rides with a negative ride_length_min are considered invalid since the trip start time cannot be greater than the trip end time
-   Rides with a ride_length_min greater than 24 hrs are considered invalid outliers for purposes of this project

```{r}
Bike_rides_2022<- Bike_rides_2022 %>%
  filter(!is.na(ride_length_min) & ride_length_min >= 1 & ride_length_min <= 1440)
```

#### **Adding columns for hour, time_of_day, day_of_week, month, and season to compare casual users and members**

#Adding column for hour of the day

This new column will contain the hour component extracted from the started_at column using the `lubridate::hour()` function.

```{r}
Bike_rides_2022$hour <- lubridate::hour(Bike_rides_2022$started_at)
```

#Adding column for time_of_day

```{r}
Bike_rides_2022 <- Bike_rides_2022 %>% mutate(time_of_day = case_when(
  hour >= 6 & hour < 9 ~ "Early Morning",
  hour >= 9 & hour < 12 ~ "Mid Morning",
  hour >= 12 & hour < 18  ~ "Afternoon",
  hour >= 18 & hour <= 23  ~ "Evening",
  hour >= 0 & hour < 3  ~ "Early Night",
  hour >= 3 & hour < 6  ~ "Late Night"))
```

#Adding a column for day_of_week the ride took place

```{r}
Bike_rides_2022<- Bike_rides_2022 %>%
  mutate(day_of_week = weekdays(started_at))
```

#Adding columns for the season and month

Function **`get_season`** is used to categorize months into seasons.

The **`sapply`** function is then used to apply the **`get_season`** function to each element in the **`month`** column.

```{r}
 get_season <- function(month) {
  if (month %in% c(12, 1, 2)) {
    return("Winter")
  } else if (month %in% c(3, 4, 5)) {
    return("Spring")
  } else if (month %in% c(6, 7, 8)) {
    return("Summer")
  } else {
    return("Fall")
  }
}

Bike_rides_2022 <- Bike_rides_2022 %>%
  mutate(month = as.integer(format(started_at, "%m")),
         season = sapply(month, get_season))
```

#### **Descriptive analysis on all rides (all figures in minutes)**

```{r}
mean(Bike_rides_2022$ride_length_min) #straight average (total ride length / rides)
median(Bike_rides_2022$ride_length_min) #midpoint number in the ascending array of ride lengths
max(Bike_rides_2022$ride_length_min) #longest ride
min(Bike_rides_2022$ride_length_min) #shortest ride
```

Key insights:

-   Average ride length is 17.27 minutes

-   Median ride length is 10.8 minutes

-   Maximum ride length is 1439.37 minutes (23.99 hours)

-   Minimum ride length is 1 min

#### **Casual rider vs the member rider**

```{r}
user_types_breakdown<-Bike_rides_2022 %>% 
  group_by(user_type) %>% 
  summarise(count = n(), Percentage = n()/nrow(Bike_rides_2022)*100)
```

``` r
# A tibble: 2 × 3
  user_type   count Percentage
  <chr>       <int>      <dbl>
1 casual    1731071       40.3
2 member    2561442       59.7
```

```{r}
ggplot(user_types_breakdown, aes(x = user_type, y = Percentage, fill = user_type)) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(x = "User Type", y = "Percentage (%)", title = "Bike Ride Distribution by User Type") +
  scale_fill_discrete(name = "User Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Key insight:

-   59.7% of riders are members, and 40.3% of riders are casual users.

#### **Descriptive analysis on casual riders and members.**

Using `aggregate()` function to compare mean, median, max and min of casual riders and members

```{r}
aggregate(Bike_rides_2022$ride_length_min ~ Bike_rides_2022$user_type, FUN = mean)
aggregate(Bike_rides_2022$ride_length_min ~ Bike_rides_2022$user_type, FUN = median)
aggregate(Bike_rides_2022$ride_length_min ~ Bike_rides_2022$user_type, FUN = max)
aggregate(Bike_rides_2022$ride_length_min ~ Bike_rides_2022$user_type, FUN = min)

```

I found out that :

-   The mean ride length for casual users and members is 24.07 min and 12.67 min, respectively.

-   The median ride length for casual users and members is 14.08 min and 9.15 min, respectively.

-   Max and min ride lengths for both type of users were nearly the same

## **Further Analysis and Visualizations:**

#### **Adding a column `ride_length_cat` for categories to get a better visual in R.**

```{r}
Bike_rides_2022 <- Bike_rides_2022 %>% 
  mutate(ride_length_cat = case_when(
   ride_length_min <11.99 ~ "< 12 min",
   ride_length_min >=12 & ride_length_min <=20.99 ~ "12-20 min",
   ride_length_min >=21 & ride_length_min <=30.99  ~ "21-30 min",
   ride_length_min >=31 & ride_length_min <=60.99  ~ "31-60 min",  ride_length_min >=60 & ride_length_min <=120.99  ~ "61-120 min",
 ride_length_min >=121 & ride_length_min <=240.99  ~ "121-240 min",
 ride_length_min >=241  ~ "241+ min"))
```

#### **Creating a data frame that will show the percentage of rides for each `user_type` within each `ride_length_cat` category.**

I will use percentage instead of raw counts.

```{r}
summary_data <-Bike_rides_2022 %>% 
  group_by(user_type, ride_length_cat) %>%
  summarise(count = n()) %>%
  group_by(user_type) %>%
  mutate(total_count = sum(count)) %>%
  mutate(percentage = (count / total_count) * 100) %>%
  select(user_type, ride_length_cat, percentage) %>%
  ungroup()
```

#### **Plot "Percentage Distribution of rides for each User Type within each Ride Length Category"**

```{r}
ggplot(summary_data, aes(x = user_type, y = percentage, fill = ride_length_cat)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "User Type", y = "Percentage (%)", title ="Percentage Distibution of rides for each User Type within each Ride Length Category") +
  scale_fill_discrete(name = "Ride Length Category") +
  theme_minimal()
```

Key insights:

-   For both user types the majority of rides fall into first two categories: "\<12 min" and "12-20 min".

-   42.44% and 24.67% of casual users ride a bike for less than 12 minutes, and between 12 and 20 minutes, respectively.

-   63.57% and 21.70% of member users ride a bike for less than 12 minutes, and between 12 and 20 minutes, respectively.

#### **Average ride length by time_of_day for members vs casual users**

```{r}
  
axis_labels <- c("Early Morning \n6am-9am", "Mid Morning \n9am-12pm", "Afternoon \n12pm-6pm", "Evening \n6pm-11pm", "Early Night \n11pm-3am", "Late Night \n3am-6am")

Bike_rides_2022 %>% 
  group_by(user_type, time_of_day) %>% 
  summarise(count = n(), average_ride_length = mean(ride_length_min)) %>% 
  ggplot(aes(x = factor(time_of_day, level = c("Early Morning", "Mid Morning", "Afternoon", "Evening", "Early Night", "Late Night")), y = average_ride_length, fill = user_type)) + 
  geom_col(position = "dodge", width = 0.4) +
  labs(x = "Time of Day", y = "Ride Length (in minutes)", title = "Average ride length by user type and time of day") +
  scale_x_discrete(labels = axis_labels) + 
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Key insights:

-   Ride length for casual riders peaks mid morning through afternoon

-   Ride length for member riders remains more steady throughout the day

#### **Average ride length by user type and hour_of_day.**

It will show us more detailed answer on when riders take longer rides during the day.

```{r}
Bike_rides_2022 %>% 
  group_by(user_type, hour) %>%
  summarise(count=n(), average_ride_length=mean(ride_length_min)) %>%
  arrange(user_type, hour) %>% 
  ggplot(aes(x=factor(hour, level= c(6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,0,1,2,3,4,5)), y=average_ride_length, fill=user_type)) + 
  geom_col(position = "dodge") +
  labs(x="Hour of day", y="Ride Length (in minutes)", title = "Average ride length by user type and hour of day")
  
```

Key insight:

-   Casual riders on average ride longer than member riders (between 25 and 30 min), peaking between 10AM - 2PM

#### **Analyzing total rides by user type and hour of the day**

```{r}
Bike_rides_2022 %>% 
  group_by(user_type, hour) %>% 
  summarise(count = n()) %>%  
  arrange(user_type, hour) %>% 
  ggplot(aes(x=factor(hour, level= c(6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,0,1,2,3,4,5)), y=count, fill=user_type)) + 
  geom_col(position = "dodge") +
  labs(x="Hour of day", y="Number of Rides", title = "Total Rides by user type and hour of day")
```

Key insights:

-   For members total rides peak during morning (7-8am) and evening (4-6 pm)

-   Casual rides also peak from 4pm to 6pm

#### **Average ride length by day_of_week for members vs casual users**

```{r}
Bike_rides_2022 %>%
  group_by(user_type, day_of_week) %>%
  summarise(mean_ride_length = mean(ride_length_min)) %>% 
  ggplot(aes(x=factor(day_of_week, level= c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), y= mean_ride_length, fill=user_type)) + 
  geom_col(position = "dodge", width = 0.4) + 
  labs(x="Day of Week", y="Ride Length (in minutes)", title = "Average ride length by user type and day of the week") +  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

Key insight:

-   Both users take longer rides on weekend

#### **Analyzing total rides by user type and season**

```{r}
Bike_rides_2022 %>% 
  group_by(user_type, season) %>% 
  summarise(count = n()) %>%  
  arrange(user_type, season) %>% 
  ggplot(aes(x=factor(season, level = c("Spring", "Summer", "Fall", "Winter")), y=count, fill=user_type)) + 
  geom_col(position = "dodge") +
  labs(x="Season", y="Number of Rides", title = "Total Rides by user type and season")
```

**Key insights:**

-   In each season we see more member rides.

-   Both riders peak in summer and decline in winter

#### **Analyzing top five starting and ending stations by user types**

**Top five starting stations for casual riders:**

Visualizing top five starting stations for casual riders. The `fct_reorder()` function sorts the data in the ascending order of value_variable

```{r}
Bike_rides_2022 %>% 
  filter(!(is.na(start_station_name))) %>% 
  filter(user_type == "casual") %>% 
  group_by(start_station_name) %>% 
  summarize(count=n()) %>% 
  arrange(-count) %>% 
  top_n(5) %>% 
  mutate(start_station_name= fct_reorder(start_station_name, count)) %>% 
  ggplot(aes(x=start_station_name, y=count, fill=count)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x="Start Station Name", y="Number of Rides", title="Top 5 starting stations for casual riders")

```

**Top 5 ending station for casual riders:**

```{r}
Bike_rides_2022 %>% 
  filter(!(is.na(end_station_name))) %>% 
  filter(user_type == "casual") %>% 
  group_by(end_station_name) %>% 
  summarize(count=n()) %>% 
  arrange(-count) %>% 
  top_n(5)
```

-   The top 5 starting and ending stations are the same for casual riders

-   The most popular starting station is Streeter Dr & Grand Ave, which is located near the park in Chicago.

**Top 5 starting stations for member riders:**

```{r}
Bike_rides_2022 %>% 
  filter(!(is.na(start_station_name))) %>% 
  filter(user_type == "member") %>% 
  group_by(start_station_name) %>% 
  summarize(count=n()) %>% 
  arrange(-count) %>% 
  top_n(5) %>% 
  mutate(start_station_name= fct_reorder(start_station_name, count)) %>% 
  ggplot(aes(x=start_station_name, y=count, fill=count)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x="Start Station Name", y="Number of Rides", title="Top 5 starting stations for member riders")
```

**Top 5 end stations for member riders:**

```{r}
Bike_rides_2022 %>% 
  filter(!(is.na(end_station_name))) %>% 
  filter(user_type == "member") %>% 
  group_by(end_station_name) %>% 
  summarize(count=n()) %>% 
  arrange(-count) %>% 
  top_n(5)
```

-   The top 5 starting and ending stations are the same for member riders except for one station

-   The top starting and ending station for member riders is Ellis Ave & 60th St which is situated within the University of Chicago

## Share:

#### Summary of key findings:

-   In general, 59.7% of riders are members, and 40.3% of riders are casual users.

-   On average, casual riders ride a bike twice longer than member riders (24.07 min and 12.67 min, respectively)

-   For both user types the majority of rides fall into first two categories: "\<12 min" and "12-20 min"

-   Ride length for casual riders peaks mid morning through afternoon, especially between 10AM - 2PM, while for members its more consistent

-   Members take more rides during morning (7-8am) and evening (4-6 pm) , casual total number of rides also peak from 4pm to 6pm

-   Both users take longer rides on weekend

-   For both user types rides peak in summer and decline in winter

### Additional data analysis to consider:

This project uses a limited public dataset due to the protection of user privacy. Ideally, the marketing team would use the full dataset to [carry out]{.underline} the following additional [analysis:]{.underline}

-   Discover which casual riders are locals and focus on their riding habits

-   Determine how well and conveniently bike stations are situated

-   Break out your single ride and day pass riders

-   Analyze the number of trips and length of trips that local casual riders take

-   Take a closer look at the riding habits of your [member]{.underline} riders

    -   Number of trips, length of trip, who's riding multiple times a day? [and]{.underline} where are they [going?]{.underline}

-   In what ways does the surrounding landscape impact the riding experience of member riders and casual riders in the area?

    -   Do they live in a dense population?

    -   Are there major employers around?

    -   Are there educational or medical institutions nearby?

    -   Get a sense of whether the surrounding infrastructure promotes riding to work, riding to school, riding to grocery stores ,eateries, shopping, coffee shops, social gatherings.

 

## Act:

#### Recommendations:

In order for the marketing team to design a new marketing strategy to convert casual riders into annual members, they need to take into account these recommendations:

-   Market around pricing for the peak hours, days of the week and season.

    Offer special rates, membership subscriptions, or loyalty programme for mid morning through afternoon hours (peak hours for casual riders), weekends (Saturday and Sunday), and for the summer months.

-   Consider which bike stations are used the most and in what areas are they situated.

    Members take the most rides during morning and evening hours, and the most popular start station is situated near the University of Chicago, which suggest that membership is used the most by university students and staff. It would be beneficial to start a marketing campaign targeting this area. For example, offer a reward for recruiting a friend to ride to work or school.

    Market to people whose home area and daily travel needs likely align well with your bike system. Focusing on dense and walkable areas, partnering with schools and companies.

    Offer special discounts, day passes around the places of leisure, nature and tourist attractions, as the analysis suggests that casual users are more prone to take a bike near parks, lakes, beaches etc.

-   Use digital media and smart advertisements:

    -   Use social media to showcase member riders who are already using bikes to go to work, school, the gym, to meet friends.

    -   Raise awareness of the ease of use, affordability, accessibility, green options and common uses for bikes rides.
